---
title: "NEON data formatting"
author: "Tad Dallas"
includes:
  in_header:
    - \usepackage{lmodern}
output:
  pdf_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    toc: yes
  html_document:
    fig_caption: yes
    fig_height: 6
    fig_width: 6
    highlight: tango
    theme: journal
---







`data.frame` with the following columns and date ranges:

+ year (2014-2019 for training data, 2020 for test data)
+ season (3 month spans 1-3, 4-6, 7-9, 10-12 month)
+ latitude and longitude
+ species richness (per site)
+ abundance (mean plot estimate per site, standardized by trapping effort)

maybe?
+ plant diversity at the site per season/year
+ annual-level variation that can capture spatial differences (e.g., mean annual temperature and mean annual precipitation, temperature seasonality, etc. for each year if available). 
+ 




Load packages. Install `ecocomDP` from CRAN if necessary
```{r}
# Load packages
# install.packages('ecocomdp')
library(ecocomDP)
library(tidyverse)

```



Use ecocomDP functions to find the mammal data set
```{r}
# how to find the mammal data in ecocomDP
data_list <- ecocomDP::search_data("mammal")
View(data_list)
```

After identifying the data package id for hte NEON small mammal data set, read it in to your R environment, and save an archive of the data as downloaded.
```{r}
# the id for the neon small mammal data mapped to ecocomDP
my_id <- "neon.ecocomdp.10072.001.001"

# read neon data -- 
#   for the two sites specified below, download size is ~52 MB. 
#   for all NEON sites, download size is ~ 595 MB
my_ecocomDP_data <- ecocomDP::read_data(
  id = my_id,
  site = c("CPER","OSBS"), #comment out this line to download all sites
  token = Sys.getenv("NEON_TOKEN"),
  check.size = FALSE)

# save data initial download locally (RDS file is only 39.7KB)
saveRDS(my_ecocomDP_data, "data/my_ecocomDP_data.RDS")
```


Read the locally save data, flatten, and calculate species richness and abundance standardized to 100 trapnights by season by year by site.
```{r}
# read in data from local file
my_ecocomDP_data <- readRDS("data/my_ecocomDP_data.RDS")


# flatten data
data_flat <- my_ecocomDP_data[[1]]$tables %>%
  ecocomDP::flatten_data() %>%
  mutate(
    season = case_when(
      month %in% c(1:3) ~ "winter",
      month %in% c(4:6) ~ "spring",
      month %in% c(7:9) ~ "summer",
      month %in% c(10:12) ~ "fall",
      TRUE ~ NA_character_)) %>%
  dplyr::select(
    observation_id, event_id,
    year, season,
    siteID, domainID, nlcdClass,
    taxon_rank,
    taxon_id, 
    variable_name, value, unit) %>%
  dplyr::filter(value > 0)

# average counts for each taxon_id within a season
data_by_site_year_season_taxon <- data_flat %>%
  group_by(siteID, year, season, taxon_id) %>%
  summarize(
    count_per_100_trapnight = mean(value)) %>%
  ungroup()
  
# sum average counts for each taxon in each season to 
# get estimate of total small mammal count per trapping effort
data_by_site_year_season <- data_by_site_year_season_taxon %>%
  dplyr::filter(count_per_100_trapnight > 0) %>%
  group_by(siteID, year, season) %>%
  summarize(
    total_count_per_100_trapnight = sum(count_per_100_trapnight),
    richness = length(unique(taxon_id)))
```


Read in NEON site information from the NEON website and merge with the small mammal table, and save in the `data/` subdirectory in this repo.
```{r}
# get site info data
site_info <- read_csv(file = "https://www.neonscience.org/sites/default/files/NEON_Field_Site_Metadata_20210226_0.csv") %>%
  dplyr::select(
    domain_id, site_id, site_name,
    site_type, 
    latitude, longitude,
    mean_elevation_m, minimum_elevation_m, maximum_elevation_m, 
    mean_annual_temperature_C, mean_annual_precipitation_mm, 
    dominant_wind_direction, 
    mean_canopy_height_m, dominant_nlcd_classes, domint_plant_species,
    usgs_geology_unit, megapit_soil_family, soil_subgroup,
    avg_number_of_green_days, avg_green_max_doy)
  

# rename columns
names(site_info) <- names(site_info) %>% gsub("field_","",.)

# combine site info data with small mammal data
data_merged <- data_by_site_year_season %>% 
  left_join(
    site_info,by = c("siteID" = "site_id"))

# write to the data subdir
write_csv(data_merged, "data/data.csv")

```































